The document title

### 2021-03-11 manuscript/manuscript.Rmd
Revise text

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) in an instrumented treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to warm up on the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 $\mathrm{km\cdot h^{-1}}$ with 1 $\mathrm{km\cdot h^{-1}}$ increment. Walking and running speeds ranged from 2 to 6 $\mathrm{km\cdot h^{-1}}$ and from 7 to 14 $\mathrm{km\cdot h^{-1}}$, respectively. Each walking speed lasted for 1 minute and each running speed for 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

@@

During the protocol, participants wore three activity monitors, one at their right hip (along the anterior axillary line, at the level of the iliac crest), another at their lower back (at the midpoint between the two posterior superior iliac spines) and the last at their right ankle (immediately superior to the lateral malleolus). Activity monitors at waist level were secured on the same elastic belt with a clip and the activity monitor placed on the ankle was secured by an elastic belt fixed with adhesive tape. In all placements the accelerometer y-axis (vertical) was aligned with the standing body longitudinal axis. All participants wore the same three activity monitors always positioned at the same places.

@@

FP and accelerometer data were processed using MATLAB (Version 8.3, Mathworks, Natick, USA), as described elsewhere [@Veras2020]. Briefly, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, and their resultant vector calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, peak ACC (pACC) values were defined as minimum height of twice the positive ACC average, during any given speed, and separated by 0.4 seconds minimum. GRF and ACC signals were adjusted through the time set by the systems clock, with manual correction based on visual inspection for possible errors, and then synchronized by the maximum cross-correlation coefficient. Peak GRF (pGRF) was determined as the highest value between 0.4 seconds before and after each pACC. The rates of change, LR for the GRF and ACC rate (AR) for the ACC, were computed through a second-order centered derivative from the beginning of the foot contact to the pGRF, and pLR and peak AR (pAR) were defined as the maximum value in the array.

@@

Regression equations to predict pGFR and pLR for each accelerometer placement (ankle, lower back and hip) and vector (resultant and vertical) were developed by linear mixed models. Body mass (kg) was tested as a predictor in all models. In addition, in the pGRF models pACC was tested as predictor, while in the pLR models pAR was tested as predictor. All predictors were tested as fixed effects and were shown to be significant. Subject and speed were tested as random effects, and the inclusion of both has showed to improve the models. Final models were chosen according to -2 log-likelihood statistics [@Field2012]. Traditional coefficient of determination ($\mathrm{R^2}$) was represented by conditional $\mathrm{R^2}$, that estimates the variance explained by the whole model [@Nakagawa2013].

@@

Models were validated using the leave-one-out cross-validation approach [@Staudenmayer2012]. For leave-one-out cross-validation each participant’s data was separated into a testing dataset (one participant at a time) with the remaining data in the training dataset. New linear mixed models with the same outcomes and predictors as determined for the entire sample were developed using the training dataset and then used to predict the outcome for the participant in the testing dataset. This process was repeated for all participants (`r max(sample_size$pGRF_N)` times). Data from the testing dataset was used in the remaining statistical analysis.

@@

Finally, our pGRF prediction equation was compared with a previously published reference equation [@Neugebauer2014], that was developed using similar methods. This compaarison was performed in three ways using: i) the whole sample; ii) a subsample of walking trials; iii) a subsample of running trials. To assess the prediction accuracy, MAE, MAPE and RMSE were calculated using pVGRF values predicted from both equations. No pLR prediction equation suited to be compared with ours was found.

@@

Regression equations were developed to predict the mechanical loading variables based on accelerometry data. In the pGRF prediction equations, body mass and pACC were significant predictors and included in the models. Table 1 shows the regression coefficients, $\mathrm{R^2}$ and accuracy indices from models derived from accelerometers placed at ankle, lower back and hip, and from the resultant vector and its vertical component. In all pGRF prediction models, $\mathrm{R^2}$ ranged from `r min(GRF_accuracy$R2_cond)` to `r max(GRF_accuracy$R2_cond)`, which shows that the predictor variables are able to explain in a great extent the outcome. Also, MAE ranged between `r myround(min(GRF_accuracy$MAE), 1)` N and `r myround(max(GRF_accuracy$MAE), 1)` N, and the MAPE from `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% to `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which is a smaller error than previously published equations [@Neugebauer2014; @Fortune2014], and similar to the error from equations previously published by out group, but which included walking only [@Veras2020]. RMSE were also computed and shown to range between `r myround(min(GRF_accuracy$RMSE), 1)` N and `r myround(max(GRF_accuracy$RMSE), 1)` N. Of all the pGRF predictions equations developed, the one based on hip-worn accelerometers and the resultant vector has shown the best performance, with an $\mathrm{R^2}$ of `r hip_GRF_accuracy$R2_cond`, and the lowest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%) and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).

@@

Afterwards, Bland-Altman plots were built to visualize the agreement between the actual and predicted pGRF and pLR, for both the resultant vector and its vertical component and all accelerometer placements. A good agreement in a Bland-Altman plot is observed when values tend to cluster around zero, showing no consistent over or underestimation of the actual values. Also, the majority of data points should lie within the limits of agreement, which are determined by \pm 1.96 standard deviation of the bias (average difference between actual and predicted values). Bland-Altman plots for the pGRF and pLR resultant vector from hip-worn accelerometers can be seen in Figure 3, and for all accelerometers placements and both resultant and vertical vectors in the Supplemental Figures S3 and S4. Values for all pGRF and pLR prediction equations developed tended to assemble around zero and between the limits of agreement. For the pGRF equations, a tendency to higher dispersion can be observed in the higher magnitudes, which correspond to the running trials, and especially in values derived from ankle-worn accelerometers (Supplemental Figure S3, Panels A and D). Prediction equations for pLR showed a more accentuated dispersion, and again in the higher magnitudes. None of the regression models presented a bias different than zero ($p > 0.05$), demonstrating that these models do not consistently over or underestimate the pGRF and pLR values. Furthermore, linear regressions were applied to identify proportional bias, that is, to test whether pGRF and pLR prediction errors were systematically influenced by their magnitude. The pGRF prediction equation based on ankle-worn accelerometers data for the resultant vector showed a significant proportional bias ($p = `r myround(coefficients(ba_regression_res_GRF$ankle)[2, 4], 2)`$), although with a low magnitude ($\mathrm{R^2} = `r myround(ba_regression_res_GRF$ankle["adj.r.squared"], 2)`$).

@@

Lastly, we compared the performance of our equations with a reference equation previously published. As the only comparable equation found was the one to predict pGRF vertical vector using hip-worn accelerometer data developed by Neugebauer et al [@Neugebauer2014], we applied it to our data and computed the MAE, MAPE and RMSE. Then, these indices were compared to the indices for our equation (Table 1). Apart from the overall equations performance, it was done a sub-analysis separating the walking and running trials. The results of these comparisons can be observed in the Table 2. Accuracy indices of our equations were all remarkably better compared to the reference equation, with a MAPE approximately 2 to 3 times lower. Furthermore, our equation performance was constant between walking and running, indicating that it can be used for both activities without loss of accuracy.

>>>>

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) on an instrumented treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to warm up on the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 $\mathrm{km\cdot h^{-1}}$ with 1 $\mathrm{km\cdot h^{-1}}$ increment. Walking and running speeds ranged from 2 to 6 $\mathrm{km\cdot h^{-1}}$ and from 7 to 14 $\mathrm{km\cdot h^{-1}}$, respectively. Each walking speed lasted for 1 minute and each running speed for 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

@@

During the protocol, participants wore three activity monitors, one at their right hip (along the anterior axillary line, at the level of the iliac crest), another at their lower back (at the midpoint between the two posterior superior iliac spines) and the last at their right ankle (immediately superior to the lateral malleolus). Activity monitors at waist level were secured on the same elastic belt with a clip and the activity monitor placed on the ankle was secured by an elastic belt fixed with adhesive tape. In all placements the accelerometer vertical axis was aligned with the standing body longitudinal axis. All participants wore the same three activity monitors always positioned at the same places.

@@

FP and accelerometer data were processed using MATLAB (Version 8.3, Mathworks, Natick, USA), as described elsewhere [@Veras2020]. Briefly, both GRF and ACC signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, and their resultant vector calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, peak ACC (pACC) values were defined as minimum height of twice the positive ACC average, during any given speed, and separated by 0.4 seconds minimum. GRF and ACC signals were adjusted through the time set by the systems clock, with manual correction based on visual inspection for possible errors, and then synchronized by the maximum cross-correlation coefficient. Peak GRF (pGRF) was determined as the highest value between 0.4 seconds before and after each pACC. The rates of change, LR for the GRF and ACC rate (AR) for the ACC, were computed through a second-order centered derivative from the beginning of the foot contact to the pGRF. The pLR and peak AR (pAR) were defined as the maximum value in the array.

@@

Regression equations to predict pGFR and pLR for each accelerometer placement (ankle, lower back and hip) and vector (resultant and vertical) were developed by linear mixed models. Body mass (kg) was tested as a predictor in all models. Additionally, pACC and pAR were tested as predictors in the pGRF and pLR models, respectively. All predictors were tested as fixed effects and were shown to be significant. Subject and speed were tested as random effects, and the inclusion of both has showed to improve the models. Final models were chosen according to -2 log-likelihood statistics [@Field2012]. Traditional coefficient of determination ($\mathrm{R^2}$) was represented by conditional $\mathrm{R^2}$, that estimates the variance explained by the whole model [@Nakagawa2013].

@@

Models were validated using the leave-one-out cross-validation approach [@Staudenmayer2012]. For leave-one-out cross-validation each participant’s data were separated into a testing dataset (one participant at a time) with the remaining data in the training dataset. New linear mixed models with the same outcomes and predictors as determined for the entire sample were developed using the training dataset and then used to predict the outcome for the participant in the testing dataset. This process was repeated for all participants (`r max(sample_size$pGRF_N)` times). Data from the testing dataset was used in the remaining statistical analysis.

@@

Finally, our pGRF prediction equation was compared with a previously published reference equation [@Neugebauer2014], that was developed using similar methods. This comparison was performed in three ways using: i) the whole sample; ii) a subsample of walking trials; iii) a subsample of running trials. To assess the prediction accuracy, MAE, MAPE and RMSE were calculated using pGRF values predicted from both equations. No pLR prediction equation suited to be compared with ours was found.

@@

Regression equations were developed to predict the mechanical loading variables based on accelerometry data. In the pGRF prediction equations, body mass and pACC were significant predictors and included in the models. Table 1 shows the regression coefficients, $\mathrm{R^2}$ and accuracy indices from models derived from accelerometers placed at ankle, lower back and hip, and from the resultant vector and its vertical component. In all pGRF prediction models, $\mathrm{R^2}$ ranged from `r min(GRF_accuracy$R2_cond)` to `r max(GRF_accuracy$R2_cond)`, which shows that the predictor variables are able to explain the variability in the outcome varibale in a great extent. Also, MAE ranged between `r myround(min(GRF_accuracy$MAE), 1)` N and `r myround(max(GRF_accuracy$MAE), 1)` N, and the MAPE from `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% to `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which is a smaller error than previously published equations [@Neugebauer2014; @Fortune2014], and similar to the error from equations previously published by our group, but which included walking only [@Veras2020]. RMSE were also computed and shown to range between `r myround(min(GRF_accuracy$RMSE), 1)` N and `r myround(max(GRF_accuracy$RMSE), 1)` N. Of all the pGRF predictions equations developed, the one based on hip-worn accelerometers and the resultant vector has shown the best performance, with an $\mathrm{R^2}$ of `r hip_GRF_accuracy$R2_cond`, and the lowest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%) and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).

@@

Afterwards, Bland-Altman plots were built to visualize the agreement between the actual and predicted pGRF and pLR, for both the resultant vector and its vertical component and all accelerometer placements. A good agreement in a Bland-Altman plot is observed when values tend to cluster around zero, showing no consistent over or underestimation of the actual values. Also, the majority of data points should lie within the limits of agreement, which are determined by \pm 1.96 standard deviation of the difference between actual and predicted values. Bland-Altman plots for the pGRF and pLR resultant vector from hip-worn accelerometers can be seen in Figure 3, and for all accelerometers placements and both resultant and vertical vectors in the Supplemental Figures S3 and S4. Values for all pGRF and pLR prediction equations developed tended to assemble around zero and between the limits of agreement. For the pGRF equations, a tendency to higher dispersion can be observed in the higher magnitudes, which correspond to the running trials, and especially in values derived from ankle-worn accelerometers (Supplemental Figure S3, Panels A and D). Prediction equations for pLR showed a more accentuated dispersion, and again in the higher magnitudes. None of the regression models presented a bias different than zero ($p > 0.05$), demonstrating that these models do not consistently over or underestimate the pGRF and pLR values. Furthermore, linear regressions were applied to identify proportional bias, that is, to test whether pGRF and pLR prediction errors were systematically influenced by their magnitude. The pGRF prediction equation based on ankle-worn accelerometers data for the resultant vector showed a significant proportional bias ($p = `r myround(coefficients(ba_regression_res_GRF$ankle)[2, 4], 2)`$), although with a low magnitude ($\mathrm{R^2} = `r myround(ba_regression_res_GRF$ankle["adj.r.squared"], 2)`$).

@@

Lastly, we compared the performance of our equations with a reference equation previously published. As the only comparable equation found was the one to predict pGRF vertical vector using hip-worn accelerometer data developed by Neugebauer et al [@Neugebauer2014], we applied it to our data and computed the MAE, MAPE and RMSE. Then, these indices were compared to the indices for our equation (Table 1). Apart from the overall equations performance, it was done a sub-analysis separating the walking and running trials. The results of these comparisons can be observed in the Table 2. Accuracy indices of our equations were all remarkably better compared to the reference equation, with a MAPE approximately 2 to 3 times lower. Furthermore, our equation performance was constant between walking and running, indicating that it can be used for both activities without loss of accuracy, which cannot be observed in the reference equation.
